Estimating duration using survival analysis

### 2023-06-02 cis-perfect-testing.tex
Move total model to appendix

The estimation of incidence from prevalence is sensitive to the tail of the duration distribution.
In \cref{ATACCC} I estimated the duration distribution from the ATACCC data, however, this study suffered from a small sample size and only 20 days of follow-up.
Therefore, the estimation of the tail of the duration distribution is driven by model assumptions.
This chapter develops methods to estimate the distribution using CIS data, which has less frequent sampling but a large sample size and long follow-up.

I take a survival analysis approach to model the coarse sampling of the CIS (see \cref{perf-test:sec:problem}).
The problem of priors can be important to the results here, due to the lack of information in the data; in addition, the prior can be used to combine the information in the CIS with the results of \cref{ATACCC} (see \cref{perf-test:sec:parameters-priors}).
Through a simulation study (see \cref{perf-test:sec:simulation-study}), I show that these methods perform well, and in this context are not sensitive to model or prior specification.
However, application in reality must account for misclassification bias, the subject of \cref{imperf-testing}.

@@

I define an individual as \emph{detectable} if they would test positive if tested, in the absence of misclassification error (which is dealt with in \cref{imperf-test}).
Episode $i$'s \emph{duration of positivity}, $D_i$ is the number of days that they are detectable for; I use time discretised to days, therefore, the duration is at least 1 day.
Note that the relationship between positivity (\ie the result of a PCR test) and infectiousness (\ie an individual's ability to pass on the virus) is complex, \todo{ref for PCR pos vs infectiousness} and beyond the scope of this thesis.
I define an episode as starting at time $B_i$ and ends at $E_i$, hence $D_i = E_i - B_i + 1$.

@@

The reason for this approach, over modelling biomarkers (as in \cref{ATACCC}) is due to two factors.
First, the large sample size in CIS means that approaches where the number of parameters scales with the number of observed infections (as it does with random effect based approaches) would be computationally challenging; this challenge is compounded by the data being stored in a trusted research environment limiting the computational power and tools available.
Second, the data is coarse with most episodes having only a single positive test; therefore, the trajectory of the viral load would not be well-identified in an individual.
Combined, these factors outweighed the downside of survival analysis only considering the binary outcome (\ie positive or negative) or the test rather than the additional information on the viral load.

The methods in this chapter deal with modelling two elements of the data-generating process, double interval censoring and truncation.
False negatives are not dealt with and left to \cref{imperf-test}.

@@

Truncation is a large problem when individuals move to the infrequent testing schedule within CIS.
Consider that, based on the estimates in \cref{ATACCC}, the median length of an episode is 14 days \todo{check this number} we would not expect many episodes to last the 28 days that can occur between tests.
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/truncation}
  \caption{Episodes data in CIS is truncated, meaning that an unknown number of infections are never detected. Consider the two individuals shown here: we collect identical data for them (a series of negative tests) yet the top individual was never infected and the bottom individual was. See the main text for why this is an issue. \label{perf-test:fig:truncation}}
\end{figure}

@@

While this chapter sets up the framework for analysing the CIS data, applying the method to CIS data requires modelling false negatives.
That is the subject of \cref{imperf-test} and hence a full description of the data and processing of it is deferred to that chapter.

@@

The individual model considers only individuals with detected episodes.
This approach is based on long-standing methodology~\autocites{heiseyModelling}{dempsterMaximum}{turnbullEmpirical}.
The conceptual approach here is to form a cohort of individuals for which we observe at least one positive test, and then consider a study which had only enrolled these individuals.
The modelling approach then corrects for this selection bias, by considering \emph{ghost} infections that occur in identical individuals but were truncated.

In contrast, the total model also takes into account data from individuals without any observed episodes.
These individuals may have truncated episodes or may not have been infected at all.
This approach utilises their test times to estimate the probability of truncated infections, ultimately providing an interpretable estimate of the number of truncated episodes in the cohort.

@@

\subsection{Individual model}\label{perf-test:sec:individual-model}

\subsubsection{Likelihood}\label{perf-test:sec:likelihood}

@@

\subsubsection{Full posterior}\label{perf-test:sec:full-posterior}

@@

\subsection{Total model}\label{perf-test:sec:total-model}

Assuming $N_\text{tot} \dist \text{NegBin}(\mu, r)$:
\begin{align}
&\sum_{n_\text{tot}=n_a}^\infty \frac{n_\text{tot}!}{(n_\text{tot}-n_a)!} p_t^{n_\text{tot}-n_a} p(n_\text{tot}) \\
&= \int \sum_{n_\text{tot}=n_a}^\infty \frac{n_\text{tot}!}{(n_\text{tot}-n_a)!} p_t^{n_\text{tot}-n_a} p(n_\text{tot} \mid \lambda) p(\lambda) d\lambda \\
&= \int \sum_{n_\text{tot}=n_a}^\infty \frac{n_\text{tot}!}{(n_\text{tot}-n_a)!} p_t^{n_\text{tot}-n_a} \frac{\lambda^n_\text{tot} e^{-\lambda}}{n_\text{tot}!} p(\lambda) d\lambda \\
&= \int \sum_{n_\text{tot}=n_a}^\infty \frac{1}{(n_\text{tot}-n_a)!} p_t^{n_\text{tot}-n_a} \lambda^{n_\text{tot}-n_a} \lambda^{n_a} e^{-\lambda} p(\lambda) d\lambda \\
&= \int \lambda^{n_a} e^{-\lambda} p(\lambda) \sum_{n_t=0}^\infty \frac{1}{n_t!} p_t^{n_t} \lambda^{n_t} d\lambda &n_t = n-n_a\\
&= \int \lambda^{n_a} e^{-\lambda} p(\lambda) e^{\lambda p_t} d\lambda \\
&= \int \lambda^{n_a} e^{-\lambda(1 - p_t)} p(\lambda) d\lambda \\
&= \int \lambda^{n_a} e^{-\lambda(1 - p_t)} \frac{b^a}{\Gamma(a)} \lambda^{a-1} e^{-b\lambda} \lambda d\lambda \\
&= \int \frac{b^a}{\Gamma(a)} \lambda^{a+n_a-1} e^{-(b+1-p_t)\lambda} \lambda d\lambda \\
&= \frac{b^a}{\Gamma(a)} \frac{\Gamma(a+n_a)}{(b+1-p_t)^{a+n_a}} \\
&\propto (b+1-p_t)^{-(a+n_a)} \\
&= (r/\mu + 1 - p_t)^{-(r+n_a)} \\
&\propto(r + \mu (1- p_t))^{-(r+n_a)}
\end{align}

@@

In this section, I consider the various parameterisation of the survival function that are available.
This has implications for the choice of prior it is not possible to be vague on all sensible parameterisations.
Informative priors might also be attractive, and allow the incorporation of the estimates from \cref{ATACCC}.

@@

An uninformative prior would be $\lambda_t \dist \Beta(\alpha, \alpha)$, with common choices of $\alpha$ being $0.5$, Jeffreys' Prior, or $1$, a flat prior.
Flat or uninformative priors on the hazard lead to very informative priors on the survival, favouring short survival times.
The sharp falloff can be seen by noting that these priors have expectation 0.5, hence $\E(S(t)) = \E \left( \prod_{i=1}^{t-1} (1-\lambda_t) \right) = 0.5^{t-1}$ which decreases very quickly.
Our prior belief is not well-represented by this prior, and in fact a priori viewed as extremely unlikely (see \cref{perf-test:fig:flat-priors}),
Therefore, I favour a prior which is weakly informative on the survival, or an informative prior used to represent our belief from the \cref{ATACCC} analysis.
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/flat-prior}
  \todo[inline]{Create flat priors figure}
  \caption{The prior predictive survival time (shown) is very short when using either a Beta(1, 1) flat prior or Beta(0.5, 0.5) uninformative prior on the hazard at each time. \label{perf-test:fig:flat-prior}}
\end{figure}

@@

The analysis in \cref{ATACCC} can be used to provide prior information for the CIS analysis.
In particular, the ATACCC study frequently samples individuals early in the infection.
Therefore, the survival distribution from that analysis should be reliable through this period.
CIS can then inform the remainder of the distribution where the previous analysis is based on extrapolation.

The model structure assumed in \cref{ATACCC} leads to positive correlation in the posterior estimates of the hazard.
It is important to propagate this correlation as part of the prior uncertainty into the present analysis.
We do this in two parts: first, approximate the previous posterior estimate of the logit-hazard as a multivariate normal (which can be viewed as an approximation of Markov melding\cref{cite Markov melding}).
Then, use a Beta process to add additional uncertainty to this.

@@

I take as the ground truth distribution for duration of positivity a combination of the estimates from \cref{ATACCC} with an inflated tail to represent what is seen within CIS.
The tail is modified based on Sarah Walker's (unpublished) work based on CIS data.
These curves and the combined curve are compared in \cref{perf-test:fig:duration-dist}.
Individual $i$'s duration of positivity, $D_i$, is then an independent draw from this distribution.

@@

\section{Discussion}

@@

>>>>

The estimation of incidence from prevalence is sensitive to the tail of the duration distribution.
In \cref{ATACCC} I estimated the duration distribution from the ATACCC data, however, this study suffered from a small sample size and only 20 days of follow-up.
\todo{Ref ATACCC chapter and in following paragraph}
Therefore, the estimation of the tail of the duration distribution is driven by model assumptions.
This chapter develops methods to estimate the distribution using CIS data, which has less frequent sampling but a large sample size and long follow-up.

I take a survival analysis approach to model the coarse sampling of the CIS (see \cref{perf-test:sec:problem}).
The problem of priors can be important to the results here, due to the lack of information in the data; in addition, the prior can be used to combine the information in the CIS with the results of \cref{ATACCC} (see \cref{perf-test:sec:parameters-priors}).
Through a simulation study (see \cref{perf-test:sec:simulation-study}), I show that these methods perform well, and in this context are not sensitive to model or prior specification.
However, application in reality must account for misclassification bias, the subject of \cref{imperf-test}.
\todo{Ref chapter incl misclassification}

@@

I define an individual as \emph{detectable} if they would test positive if tested, in the absence of misclassification error (which is dealt with in \cref{imperf-test}).
\todo{Ref chapter incl misclassification}
Episode $i$'s \emph{duration of positivity}, $D_i$ is the number of days that they are detectable for; I use time discretised to days, therefore, the duration is at least 1 day.
Note that the relationship between positivity (\ie the result of a PCR test) and infectiousness (\ie an individual's ability to pass on the virus) is complex, \todo{ref for PCR pos vs infectiousness} and beyond the scope of this thesis.
I define an episode as starting at time $B_i$ and ends at $E_i$, hence $D_i = E_i - B_i + 1$.

@@

The reason for this approach, over modelling biomarkers (as in \cref{ATACCC}) is due to two factors.
\todo{Ref ATACCC chapter}
First, the large sample size in CIS means that approaches where the number of parameters scales with the number of observed infections (as it does with random effect based approaches) would be computationally challenging; this challenge is compounded by the data being stored in a trusted research environment limiting the computational power and tools available.
Second, the data is coarse with most episodes having only a single positive test; therefore, the trajectory of the viral load would not be well-identified in an individual.
Combined, these factors outweighed the downside of survival analysis only considering the binary outcome (\ie positive or negative) or the test rather than the additional information on the viral load.

The methods in this chapter deal with modelling two elements of the data-generating process, double interval censoring and truncation.
False negatives are not dealt with and left to \cref{imperf-test}.
\todo{Ref chapter incl misclassification}

@@

Truncation is a large problem when individuals move to the infrequent testing schedule within CIS.
Consider that, based on the estimates in \cref{ATACCC}, the median length of an episode is 14 days \todo{check this number} we would not expect many episodes to last the 28 days that can occur between tests.
\todo{Ref ATACCC chapter}
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/truncation}
  \caption{Episodes data in CIS is truncated, meaning that an unknown number of infections are never detected. Consider the two individuals shown here: we collect identical data for them (a series of negative tests) yet the top individual was never infected and the bottom individual was. See the main text for why this is an issue. \label{perf-test:fig:truncation}}
\end{figure}

@@

While this chapter sets up the framework for analysing the CIS data, applying the method to CIS data requires modelling false negatives.
That is the subject of \cref{imperf-test} and hence a full description of the data and processing of it is deferred to that chapter.
\todo{Ref chapter incl misclassification}

@@

I model only individuals with detected episodes.
This approach is based on long-standing methodology~\autocites{heiseyModelling}{dempsterMaximum}{turnbullEmpirical}.
The conceptual approach here is to form a cohort of individuals for which we observe at least one positive test, and then consider a study which had only enrolled these individuals.
The modelling approach then corrects for this selection bias, by considering \emph{ghost} infections that occur in identical individuals but were truncated.

I have also considered a model including individuals without any observed episodes (see \cref{appendix:total-model} for details).
These individuals may have truncated episodes or may not have been infected at all.
However, I found that this model has little difference in results while increasing computational complexity and requiring assumptions on reinfection (discussed in \cref{perf-test:sec:discussion}).

@@

\subsection{Likelihood}\label{perf-test:sec:likelihood}

@@

\subsection{Full posterior}\label{perf-test:sec:full-posterior}

@@

@@

In this section, I consider the various parameterisation of the survival function that are available.
This has implications for the choice of prior it is not possible to be vague on all sensible parameterisations.
Informative priors might also be attractive, and allow the incorporation of the estimates from \cref{ATACCC}.
\todo{Ref ATACCC chapter}

@@

An uninformative prior would be $\lambda_t \dist \Beta(\alpha, \alpha)$, with common choices of $\alpha$ being $0.5$, Jeffreys' Prior, or $1$, a flat prior.
Flat or uninformative priors on the hazard lead to very informative priors on the survival, favouring short survival times.
The sharp falloff can be seen by noting that these priors have expectation 0.5, hence $\E(S(t)) = \E \left( \prod_{i=1}^{t-1} (1-\lambda_t) \right) = 0.5^{t-1}$ which decreases very quickly.
Our prior belief is not well-represented by this prior, and in fact a priori viewed as extremely unlikely (see \cref{perf-test:fig:flat-priors}),
Therefore, I favour a prior which is weakly informative on the survival, or an informative prior used to represent our belief from the \cref{ATACCC} analysis.
\todo{Ref ATACCC chapter}
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/flat-prior}
  \todo[inline]{Create flat priors figure}
  \caption{The prior predictive survival time (shown) is very short when using either a Beta(1, 1) flat prior or Beta(0.5, 0.5) uninformative prior on the hazard at each time. \label{perf-test:fig:flat-prior}}
\end{figure}

@@

The analysis in \cref{ATACCC} can be used to provide prior information for the CIS analysis.
\todo{Ref ATACCC chapter here and next few}
In particular, the ATACCC study frequently samples individuals early in the infection.
Therefore, the survival distribution from that analysis should be reliable through this period.
CIS can then inform the remainder of the distribution where the previous analysis is based on extrapolation.

The model structure assumed in \cref{ATACCC} leads to positive correlation in the posterior estimates of the hazard.
\todo{Ref ATACCC chapter}
It is important to propagate this correlation as part of the prior uncertainty into the present analysis.
We do this in two parts: first, approximate the previous posterior estimate of the logit-hazard as a multivariate normal (which can be viewed as an approximation of Markov melding~\autoref{goudieJoining}).
Then, use a Beta process to add additional uncertainty to this.

@@

I take as the ground truth distribution for duration of positivity a combination of the estimates from \cref{ATACCC} with an inflated tail to represent what is seen within CIS.
\todo{Ref ATACCC chapter}
The tail is modified based on Sarah Walker's (unpublished) work based on CIS data.
These curves and the combined curve are compared in \cref{perf-test:fig:duration-dist}.
Individual $i$'s duration of positivity, $D_i$, is then an independent draw from this distribution.

@@

\section{Discussion} \label{perf-test:sec:discussion}

@@

\appendix

\chapter{Including all CIS individuals in survival analysis}\label{perf-test:sec:total-model} \label{appendix:total-model}

Assuming $N_\text{tot} \dist \NegBin(\mu, r)$:
\begin{align}
&\sum_{n_\text{tot}=n_a}^\infty \frac{n_\text{tot}!}{(n_\text{tot}-n_a)!} p_t^{n_\text{tot}-n_a} p(n_\text{tot}) \\
&= \int \sum_{n_\text{tot}=n_a}^\infty \frac{n_\text{tot}!}{(n_\text{tot}-n_a)!} p_t^{n_\text{tot}-n_a} p(n_\text{tot} \mid \lambda) p(\lambda) d\lambda \\
&= \int \sum_{n_\text{tot}=n_a}^\infty \frac{n_\text{tot}!}{(n_\text{tot}-n_a)!} p_t^{n_\text{tot}-n_a} \frac{\lambda^n_\text{tot} e^{-\lambda}}{n_\text{tot}!} p(\lambda) d\lambda \\
&= \int \sum_{n_\text{tot}=n_a}^\infty \frac{1}{(n_\text{tot}-n_a)!} p_t^{n_\text{tot}-n_a} \lambda^{n_\text{tot}-n_a} \lambda^{n_a} e^{-\lambda} p(\lambda) d\lambda \\
&= \int \lambda^{n_a} e^{-\lambda} p(\lambda) \sum_{n_t=0}^\infty \frac{1}{n_t!} p_t^{n_t} \lambda^{n_t} d\lambda &n_t = n-n_a\\
&= \int \lambda^{n_a} e^{-\lambda} p(\lambda) e^{\lambda p_t} d\lambda \\
&= \int \lambda^{n_a} e^{-\lambda(1 - p_t)} p(\lambda) d\lambda \\
&= \int \lambda^{n_a} e^{-\lambda(1 - p_t)} \frac{b^a}{\Gamma(a)} \lambda^{a-1} e^{-b\lambda} \lambda d\lambda \\
&= \int \frac{b^a}{\Gamma(a)} \lambda^{a+n_a-1} e^{-(b+1-p_t)\lambda} \lambda d\lambda \\
&= \frac{b^a}{\Gamma(a)} \frac{\Gamma(a+n_a)}{(b+1-p_t)^{a+n_a}} \\
&\propto (b+1-p_t)^{-(a+n_a)} \\
&= (r/\mu + 1 - p_t)^{-(r+n_a)} \\
&\propto(r + \mu (1- p_t))^{-(r+n_a)}
\end{align}
